ropenblas: Download, Compile and Link OpenBLAS Library with R

### 2021-03-09 paper/paper.Rmd
Update paper
Update

The `ropenblas` package aims to facilitate the day-to-day life of R programmers who want a little more performance on GNU/Linux systems, without removing the possibility that specific configurations are made, if they deem convenient. Through the package's `ropenblas()` and `rcompiler()` functions, the library user will be able to compile and link the R language in his GNU/Linux distribution with the OpenBLAS library, all within R and in a very simple. All functions work without being influenced by the GNU/Linux distribution and are independent of their repositories, that is, it does not matter which GNU/Linux distribution is being used. Linking the OpenBLAS library to R will bring better computational performance to the language in the most diverse algebraic operations commonly used in areas such as statistics, data science and machine learning.

@@

The `ropenblas` package aims to allow algebraic computing of R to be performed using the OpenBLAS library, that is, it allows easy linking of R to the OpenBLAS library on GNU/Linux systems without depending on the distribution repositories. This will allow several researches in the areas of statistics, data science and machine learning to take advantage of a more efficient performance in algebraic calculations, for example, multiplication, factorization and matrix inversion. The `ropenblas` library, version 0.2.9 will also allow the R programmer to have, in his GNU/Linux distribution, several compiled versions of the R language giving the possibility to easily switch between these versions, this being an Open Source functionality that is only possible in some commercial IDEs of R. All this is done within the R language, minimizing the chance of less experienced users to break their operating system by running several instructions that are not perfectly understood.

The fact that the `ropenblas` package does not depend on the repositories of the GNU/Linux distribution will allow that in more stable distributions the R programmer will have at his disposal the most recent version of the OpenBLAS libraries and the R language. Everything is done in a safe way, a since the `ropenblas` package uses the stable versions of the official development repositories of the OpenBLAS library and the R programming language, respectively. Until the present version, the package has more than 1000 downloads, having an average of more than 1000 downloads in the month prior to the date of this submission, based on the official R language repositories.

@@

The `ropenblas` package is already available on the Comprehensive R Archive Network - CRAN, currently in version 0.2.9 and the project is maintained on GitHub at <https://github.com/prdm0/ropenblas> where contributors can find others details of the code, information, as well as being able to contribute with the development of the project. Information can also be found on the project website. On the website it is also possible to read the `NEWS.md` file with details of the versions and the focus of the current development. The site is deposited at <https://prdm0.github.io/ropenblas/>. Suggestions for improvements and bug reports can be sent via the link <https://github.com/prdm0/ropenblas/issues>. You can find out how to contribute to the package by accessing the `CONTRIBUTING.md` file at <https://github.com/prdm0/ropenblas/blob/master/CONTRIBUTING.md>

@@

The `ropenblas` library exports six functions for use which are the `rcompiler()`, `ropenblas()`, `last_version_r()`, `last_version_openblas()`, `link_again()` and `rnews()`. All of them are very simple to use and have few arguments that are sufficient to maintain flexibility of use. Any example that follows will consider that the installation of the `ropenblas` package has been carried out and the package has been loaded (`library(ropenblas)`). In addition, functions like `rcompiler()` and `ropenblas()` do not return content or data structures that are of any practical use. What these functions do is configure the GNU/Linux system to use R, configure different versions of the language, switch between versions and link with the OpenBLAS library. It is also possible to obtain a summary of the versions of R and the OpenBLAS library that are available.

@@

The function `last_version_r()` automatically searches, in the official repositories of language R, for information about versions of language R. Its general use is `last_version_r(major = NULL)` , where the argument `major` indicates which is the largest version of R that you want to search for the version list. Therefore, for the argument `major` a number must be passed, preferably an integer that indicates which is the largest version to be considered.

@@

The `last_version_openblas()` function works similarly to the `last_version_r()` function, returning a list of three elements named in the same way with the information from the latest version, all the versions and the number of versions of the OpenBLAS library, respectively.

@@

This function is responsible for compiling a version of the R language. The x argument is the version of R that you want to compile. For example, `x = "4.0.4"` will compile and link R-4.0.4 version as the major version on your system. By default (`x = NULL`) will be compiled the latest stable version of the R. For example, to compile the latest stable version of the R language, run `rcompiler()`. The `rcompiler()` function can only be used if the user is an administrator of the GNU/Linux distribution. If the user is using programming IDE's, a screen similar to the image below will be displayed requesting the entry of the system administrator password.

@@

If the goal is to install the R language, switch between versions of R, and link the installed versions of the language with the OpenBLAS library, you shouldn't have to worry about the `with_blas` and `complementary_flags` arguments, respectively. These arguments are useful for a minority of programmers who feel very specific needs to pass complementary flags to be considered in the R language compilation process, inclusion of complementary library directories, among other arguments that can be found in the official language manuals. By default, if nothing is passed to the `with_blas` and `complementary_flags` arguments, the compilation will be performed as follows:

@@

-   `x`: String with the version of the OpenBLAS library to be compiled, installed and linked with the main R installation (by default it is considered the latest version);
-   `restart_r`: Logical value (default `restart_r = TRUE`) to update the R section after compiling, installing and linking the OpenBLAS library.

@@

The code below is a small example of the benefits of considering linking the R language to the OpenBLAS library, in which singular-value decomposition of a rectangular matrix is computed, codes that are executed on the same machine, version of R and section. The first part of the code was executed with R without being linked to a version of the OpenBLAS library and the second part was executed after using the `ropenblas()` function. Note that the code when executed in R linked to a version of the OpenBLAS library can be clearly more efficient:

@@

Through a banchmark it is possible to better understand the performance gain that can be achieved by linking the R language to the OpenBLAS library. Figure 2 presents the banchmarks in the form of violin plot, in which 100 reproductions of the `svd(X)` expression were considered, in the form of the code above, with the R linked to the BLAS library and linked to the OpenBLAS library, respectively, on the same hardware. It was observed that the average time of execution of the routine `svd(X)` considering the OpenBLAS library was less than 10 times the time necessary to execute it in R linking to a non-optimized version of BLAS, being the average time of 0.64 and 6.52 seconds, respectively.

@@

The `link_again()` function links again the OpenBLAS library with the R language, being useful to correct problems of untying the OpenBLAS library that is common when the operating system is updated. The function be able to link again the R language with the OpenBLAS library.

Thus, `link_again()` will only make the relinkagem when in some previous section of R the `ropenblas()` function has been used for the initial binding of the R language with the OpenBLAS library.

The use of the function is quite simple, just by running the code `link_again()`, since the function has no arguments. It will automatically detect if there was a link break that will be rebuilt again without the need for any compilation. From time to time, after a major update of the operating system, it may be convenient to run the `link_again()` function. Link breakage rarely occurs, but if it does, it can be resolved quickly. The following code and image exemplify a possible reconstruction of symbolic links using the `link_again()` function:

>>>>

The `ropenblas` package aims to facilitate the day-to-day life of R programmers who want a little more performance on GNU/Linux systems, without removing the possibility that specific configurations are made, if they deem convenient. Through the package's `ropenblas()` and `rcompiler()` functions, the library user will be able to compile and link the R language in his GNU/Linux distribution with the OpenBLAS library, all within R and in a very simple. All functions work without being influenced by the GNU/Linux distribution and are independent of their repositories, that is, it does not matter which GNU/Linux distribution is being used. Linking the OpenBLAS library to R will bring better computational performance to the language in the most diverse algebraic operations commonly used in areas such as statistics, data science, and machine learning.

@@

The `ropenblas` package aims to allow algebraic computing of R to be performed using the OpenBLAS library, that is, it allows easy linking of R to the OpenBLAS library on GNU/Linux systems without depending on the distribution repositories. This will allow several researches in the areas of statistics, data science, and machine learning to take advantage of more efficient performance in algebraic calculations, for example, multiplication, factorization, and matrix inversion. The `ropenblas` library, version 0.2.9 will also allow the R programmer to have, in his GNU/Linux distribution, several compiled versions of the R language giving the possibility to easily switch between these versions, this being an Open Source functionality that is only possible in some commercial IDEs of R. All this is done within the R language, minimizing the chance of less experienced users to break their operating system by running several instructions that are not perfectly understood.

The fact that the `ropenblas` package does not depend on the repositories of the GNU/Linux distribution will allow that in more stable distributions the R programmer will have at his disposal the most recent version of the OpenBLAS libraries and the R language. Everything is done safely, since the `ropenblas` package uses the stable versions of the official development repositories of the OpenBLAS library and the R programming language, respectively. Until the present version, the package has more than 1000 downloads, having an average of more than 1000 downloads in the month before the date of this submission, based on the official R language repositories.

@@

The `ropenblas` package is already available on the Comprehensive R Archive Network - CRAN, currently in version 0.2.9, and the project is maintained on GitHub at <https://github.com/prdm0/ropenblas> where contributors can find other details of the code, information, as well as being able to contribute with the development of the project. Information can also be found on the project website. On the website, it is also possible to read the `NEWS.md` file with details of the versions and the focus of the current development. The site is deposited at <https://prdm0.github.io/ropenblas/>. Suggestions for improvements and bug reports can be sent via the link <https://github.com/prdm0/ropenblas/issues>. You can find out how to contribute to the package by accessing the `CONTRIBUTING.md` file at <https://github.com/prdm0/ropenblas/blob/master/CONTRIBUTING.md>.

@@

The `ropenblas` library exports six functions for use which are the `rcompiler()`, `ropenblas()`, `last_version_r()`, `last_version_openblas()`, `link_again()` and `rnews()`. All of them are very simple to use and have few arguments that are sufficient to maintain the flexibility of use. Any example that follows will consider that the installation of the `ropenblas` package has been carried out and the package has been loaded (`library(ropenblas)`). Also, functions like `rcompiler()` and `ropenblas()` do not return content or data structures that are of any practical use. What these functions do is configure the GNU/Linux system to use R, configure different versions of the language, switch between versions, and link with the OpenBLAS library. It is also possible to obtain a summary of the versions of R and the OpenBLAS library that are available.

@@

The function `last_version_r()` automatically searches, in the official repositories of language R, for information about versions of language R. Its general use is `last_version_r(major = NULL)`, where the argument major indicates which is the largest version of R that you want to search for the version list. Therefore, for argument `major`, a number must be passed, preferably an integer that indicates which is the largest version to be considered.

@@

The `last_version_openblas()` function works similarly to the `last_version_r()` function, returning a list of three elements named in the same way with the information from the latest version, all the versions, and the number of versions of the OpenBLAS library, respectively.

@@

This function is responsible for compiling a version of the R language. The x argument is the version of R that you want to compile. For example, `x = "4.0.4"` will compile and link the R-4.0.4 version as the major version on your system. By default (`x = NULL`) will be compiled the latest stable version of the R. For example, to compile the latest stable version of the R language, run `rcompiler()`. The `rcompiler()` function can only be used if the user is an administrator of the GNU/Linux distribution. If the user is using programming IDE's, a screen similar to the image below will be displayed requesting the entry of the system administrator password.

@@

If the goal is to install the R language, switch between versions of R, and link the installed versions of the language with the OpenBLAS library, you shouldn't have to worry about the `with_blas` and `complementary_flags` arguments, respectively. These arguments are useful for a minority of programmers who feel very specific needs to pass complementary flags to be considered in the R language compilation process, the inclusion of complementary library directories, among other arguments that can be found in the official language manuals. By default, if nothing is passed to the `with_blas` and `complementary_flags` arguments, the compilation will be performed as follows:

@@

-   `x`: String with the version of the OpenBLAS library to be compiled, installed, and linked with the main R installation (by default it is considered the latest version);
-   `restart_r`: Logical value (default `restart_r = TRUE`) to update the R section after compiling, installing, and linking the OpenBLAS library.

@@

The code below is a small example of the benefits of considering linking the R language to the OpenBLAS library, in which singular-value decomposition of a rectangular matrix is computed, codes that are executed on the same machine, version of R, and section. The first part of the code was executed with R without being linked to a version of the OpenBLAS library and the second part was executed after using the `ropenblas()` function. Note that the code when executed in R linked to a version of the OpenBLAS library can be more efficient:

@@

Through a benchmark it is possible to better understand the performance gain that can be achieved by linking the R language to the OpenBLAS library. Figure 2 presents the benchmarks in the form of a violin plot, in which 100 reproductions of the `svd(X)` expression were considered, in the form of the code above, with the R linked to the BLAS library and linked to the OpenBLAS library, respectively, on the same hardware. It was observed that the average time of execution of the routine `svd(X)` considering the OpenBLAS library was less than 10 times the time necessary to execute it in R linking to a non-optimized version of BLAS, being the average time of 0.64 and 6.52 seconds, respectively.

@@

The `link_again()` function links again the OpenBLAS library with the R language, being useful to correct problems of untying the OpenBLAS library that is common when the operating system is updated. The function can link again the R language with the OpenBLAS library.

Thus, `link_again()` will only make the linkage when in some previous section of R the `ropenblas()` function has been used for the initial binding of the R language with the OpenBLAS library.

The use of the function is quite simple, just by running the code `link_again()` since the function has no arguments. It will automatically detect if there was a link break that will be rebuilt again without the need for any compilation. From time to time, after a major update of the operating system, it may be convenient to run the `link_again()` function. Link breakage rarely occurs, but if it does, it can be resolved quickly. The following code and image exemplify a possible reconstruction of symbolic links using the `link_again()` function:
